%!TEX root = ../thesis.tex
%*******************************************************************************
%****************************** Third Chapter **********************************
%*******************************************************************************
\chapter{Outlook}

% **************************** Define Graphics Path **************************
\ifpdf
\graphicspath{{Chapter5/Figs/Raster/}{Chapter5/Figs/PDF/}{Chapter5/Figs/}}
\else
\graphicspath{{Chapter5/Figs/Vector/}{Chapter5/Figs/}}
\fi

\section{Significance of results}
% Few single cells at apex maintained
Our analysis presented herein has given extensive novel insights into the dynamics of
the stem cell niche of AT. In particular, we have showed how the SAM appears to
maintain a few single cells that are truly expressing CLV3, and thus also set
the grounds for further investigations into the variance in and extent of this small
pool of cells; for example, questions we have not addressed herein are the
variability in the size of the true CLV3 domain in relationship to the overall
size and shape of the SAM itself. 

% Dynamic perspective
While previous research has primarily focused on the static view of the stem
cell niche, we have here provided a description of the dynamic component --
something which has previously been severely lacking. Because of this, we are
able to better investigate the hypothesis of epidermal regulation in the
CZ, and have also pointed at cues of the extent of differentiation driving peripheral
signals. That is, our observed differences in functional behaviour for cells at
various distances from the CZ suggests a possible domain for where peripheral
signalling is either present, or simply begins to affect cell functionality.
This information can be of use in particular for better validating computational
models, where
understanding the extent of unknown regulatory mechanisms can aid in unravelling
their possible nature and from there guide experiments.

% 
We have also shown how the CLV3 apex does not coincide with the geometrical
apex, something which might have an impact on anisotropic growth and
phyllotaxis. This is a completely unprecedented result, which might provide
future insights into how the interplay between signalling in the CZ and PZ is orchestrated
in order for plants to undergo primordia initiation.

\section{Expansion of modelling framework}
The models introduced to support hypotheses on regulation are very simplistic,
which while sufficient for  
our purposes does not capture the complexity and intricacies of the real tissue.
Modern software tools are now able to generate contact maps from raw confocal
images, thus creating the means for accurately comparing model simulations to
the real tissue. This is a possible extension to the work herein, which would
also be a natural future validation step. 

Establishing models for the higher dimensions would in addition allow for validating
models relating to the combined aspect of our results. For example, is epidermal
regulation sufficient for producing both our distribution for the CLV3
intensities with respect to the distance to the apex, as well as the apparent
distributional shift between the L1 and L2? A simple model in e.g.\ 2D could
likely verify or reject such an hypothesis.

Also a spatial model relating CLV3 expression to growth rates could be established to test if
there are mechanical constraints limiting the growth rate of the apical cells in
comparison of their neighbours, and possible biochemical signals regulating
this. The combined mechanical and biochemical framework is likely also something
which will be able to provide a more complete picture of meristem growth,
relating to both scaling of the stem cell niche itself, but also how that
relates to cell features such as growth and division rates depending on their
spatial localisation.

\section{Data investigation}
% Pin1
A natural and very feasible next step in quantifying the dynamics of the genes
present in the SAM would be to also include the PIN1 reporter, which is indicative of
auxin transport, and observe how it behaves in relation to the CLV3 signal.
Possibly this would aid in elucidating
the extent of both technical noise and what observed fluctuations are true
biological events. It also allows insight into the auxin transport
perspective of meristem growth and CLV3 expression, and would resolve to what
extent auxin transport is indeed restored over time. it would further provide
information on how this affects the periodicity we observe for the CLV3 domain.

% Different layers
There are multiple additional aspects of the data analysis not fully
investigated in this thesis. This includes more thorough research into
differences between the different layers of the SAM, and how the spatial
localisation relates to e.g.\ cell growth. In particularly the CZ of the SAM,
steady maintenance between the highest expressing cells and their contact areas
to neighbouring cells could prove to gain a better understanding of
the extent and regulation of the CLV3 expressing cells. 

The possible clustering of long-lived cells in the deeper, central regions
tending towards the OC is somewhat
hampered by the difficulty to accurately identify cells within the tissue.
One possibility in this case would be to develop methods able to determine
cellular age from static images, where the plant can then be cut in the coronal
plane in order to improve imaging. This would be enabled by information about, among
other measures, relative cell wall angles between neighbouring cells. An
additional approach could be to tag cells with cell cycle identifying markers,
although also in this case it is likely to be hampered by the penetration depth
of imaging tools.

As the data in addition includes multiple segmentation errors of various types,
a re-evaluation of these could help in better narrowing down the behaviour of
especially individual cell lines -- something which in our analysis is slightly
hampered due to lesser tracking quality outside of the CZ. This holds especially
true for the individual timepoints in the tracking which have been manually
corrected for, and therefore do not include cells outside of the CZ; a simple
re-tracking of these timepoints would provide both more data for the data set
overall, but would in particular enable insight into the cell line dynamics of
the subepidermis, and how they correspond to the ones in the L1. Performing an
accurate segmentation step also for plant 18 would on top of this provide a
larger data set as basis for extended analyses.

\section{New experiments}
% WUS or other tracker
Because of the difficulty in unravelling which parts of a GRN is essential for
giving it its correct functionality, it is also difficult to know which
molecules to track for in \textit{in vivo} time-lapse assays.
In the analysis of the regulation of the stem cell niche, having access to
tracker information also for the WUS would improve the identification of
core network dynamics significantly. Extending the network further, to be able
to track also the patterning activity of auxin on top of the PIN1 data give useful
information not only of when and how much auxin transport is activated, but also
where this localisation takes place in the tissue. Being able to relate this
information in particular to the behaviour of cells in particular lineages would
provide information of how the CLV3 and cell dynamics depend on auxin localisation.

% Division driving gene?
As the CLV3 level of expression does not appear to provide cues for when a cell
is to undergo a division event, identifying the core genes involved in stem cell
maintenance and how these interact with the cell proliferation pathway is a
future endeavour. In the event that no such genes exist in the direct sense,  
and that predominantly mechanical signalling instead is what activates
proliferation genes, the interplay between the mechanical and the chemical is
perhaps the more relevant. The question in that case would then primarily be to identify
growth-inducing agents, rather than direct division event markers. Both auxin
and cytokinin are two such examples,  where they are known to regulate growth in the
SAM, but where the quantification on the single cell level is missing.

% Wildtype
A drawback with the analysis presented herein in the reliance on meristems grown
on NPA for the analysis which is better suited for the steady state. However,
the alternative is to physically remove organs and primordia to get access
to imaging the meristem, which is a major source of perturbations in itself.
Developing protocols for performing this type of research would  
greatly enhance the quality and robustness by which the analysis of e.g.\ the
CLV3 distributions and behaviour over time is done. A perhaps simpler
experimental setup at the current stage would be to grow plantlings initially on
NPA, and in a controlled manner transferring them to a non-NPA medium. This
would allow for tracking of the system when it is known that auxin transport is
being reestablished in the meristem. Performing this experiment on multiple
plants in synchrony would reduce the technical variation massively in comparison
to the current data.

% CLV3 peptide
Lastly, the reason for the use of a promoter tagging molecule for CLV3 
is the size of the CLV3 peptide. As attempts to track the protein
itself leads to non-functional plant phenotypes, understanding the specific
protein patterning is as of yet practically unfeasible in a dynamic setting. The
future might hold advances in methodology that enables direct feedback of
peptide localisation, which undoubtedly would provide a major stepping stone
towards truly understanding plant development. A more feasible first step would
nevertheless be to perform the same type of experiment with a less stable
reporter in order to attain a better grasp of CLV3, and thus stem cell, dynamics.

